Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-26 appendix.tex
Beginning to read appendix.

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
\[
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
\]
and the partial log-likelihood
\[
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
\]
so that \eqref{eq:site_pattern_profile_likelihood_mean} is
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
\hline
\end{tabular}
\caption{
Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values multiplied by $1/32$.
Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
There are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
}
\label{tab:farris_likelihoods}
\end{table}

@@

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
We show that the likelihood remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t)$ we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
An analogous statement holds for $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.

@@

$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
Likelihood calculations on the InvFels tree $\tau_1$.
Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
There are $3^4=81$ possible forms for the likelihood.
}
\label{tab:likelihoods}
\end{table}

@@

\begin{proof}
We obtain
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t})
\]
as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
\[
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
\]
and
\[
\tilde{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\]
By straightforward calculus, the function $f(x)=a_1\log(1+x)+a_2\log(1-x)$ attains a maximum at $\hat{x} = (a_1-a_2)/(a_1+a_2)$, and so we have a closed form for each component of $\tilde{t}$.
Using this estimate, we have the lower bound
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\tilde{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t}).
\]
We compute the lower bound $C_0(x^*,y^*)$ above given $\emptyset$ as the maximal ancestral state split for all site splits and $\tau=\tau_1$.
We need only consider the remaining $u=1,\ldots,80$ possibilities for choices of ancestral state splits in Table~\ref{tab:likelihoods} upon excluding cases of infinite branch lengths (i.e., any of $x_1,y_1,x_2,y_2,w$ equal to zero) and the redundant cases of $x_1 > x_2$ and $y_1 > y_2$.
We compute upper bounds $C_{1,u}(x^*,y^*)$ and show in Fig.~\ref{fig:max-anc-state} the region where $C_0(x^*,y^*) > \max_u C_{1,u}(x^*,y^*)$.
\end{proof}

>>>>

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
\[
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
\]
%EM Curious: is this an established use of the term "partial"?
and the partial log-likelihood
\[
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
\]
so that \eqref{eq:site_pattern_profile_likelihood_mean} is
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
\hline
\end{tabular}
\caption{
Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values multiplied by $1/32$.
Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
}
\label{tab:farris_likelihoods}
\end{table}

@@

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
We show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.

%EM It seems to me that the point here is that these swaps in the generating probabilities are matched by swaps in the likelihoods, meaning that things stay the same. A topic sentence to this effect would help if that's right.
The corresponding possibilities for the likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.
The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
Exchanging $x_1$ and $x_2$ also does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t)$.

Thus, we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
An analogous statement holds for $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.

@@

$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
Likelihood calculations on the InvFels tree $\tau_1$.
Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the likelihood.
}
\label{tab:likelihoods}
\end{table}

@@

\begin{proof}
Recall
\[
\ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\]
We obtain
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t})
\]
as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
\[
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
\]
and
\[
\tilde{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\]
By straightforward calculus, the function $f(x)=a_1\log(1+x)+a_2\log(1-x)$ attains a maximum at $\hat{x} = (a_1-a_2)/(a_1+a_2)$, and so we have a closed form for each component of $\tilde{t}$.
Using this estimate, we have the lower bound
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\tilde{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t}).
\]
%EM C_0 needs a better definition here. Too much going on.
We compute the lower bound $C_0(x^*,y^*)$ above given $\emptyset$ as the maximal ancestral state split for all site splits and $\tau=\tau_1$.
We need only consider the remaining $u=1,\ldots,80$ possibilities for choices of ancestral state splits in Table~\ref{tab:likelihoods} upon excluding cases of infinite branch lengths (i.e., any of $x_1,y_1,x_2,y_2,w$ equal to zero) and the redundant cases of $x_1 > x_2$ and $y_1 > y_2$.
%EM Also suggest a real def for C_{1,u}.
We compute upper bounds $C_{1,u}(x^*,y^*)$ and show in Fig.~\ref{fig:max-anc-state} the region where $C_0(x^*,y^*) > \max_u C_{1,u}(x^*,y^*)$.
\end{proof}
